---
title: "Assignment 3"
output:
  html_document: default
  pdf_document: default
date: "2024-09-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#imports
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("readr")
install.packages("dplyr")
library(readr)
library(dplyr)
```

## Exercise 1

```{r}
#Load the CSV file from URL
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv"
majors_df <- read_csv(url)

#Filtering dataset
filtered_majors <- majors_df %>%
  filter(grepl("DATA|STATISTICS", Major, ignore.case = TRUE))

print(filtered_majors)
```

## Exercise 2

```{r}
fruits <- c("bell pepper", "bilberry", "blackberry", "blood orange", 
            "blueberry", "cantaloupe", "chili pepper", "cloudberry", 
            "elderberry", "lime", "lychee", "mulberry", 
            "olive", "salal berry")

# Display the result
print(fruits)
cat("c(", paste(shQuote(fruits, type = "cmd"), collapse = ", "), ")", sep = "")
```

## Exercise 3

```         
(.)\1\1
```

This expression matches a single character that repeats three times consecutively. For instance, "aaa".

```         
"(.)(.)\2\1"
```

This expression matches two characters (. and .) followed by the same characters in reverse order. For instance, it would match "xyyx"

```         
(..)\1
```

This matches two consecutive characters (..) that are repeated in the same order. For instance, "1212"

```         
"(.).\1.\1"
```

This expression matches a single character (.), followed by any character, then the first character, followed by another any character, and then the same first character again. For instance, "aba"

```         
"(.)(.)(.).*\3\2\1"
```

This expression matches three distinct characters (.)(.)(.) followed by any number of characters (.\*), and then the same three characters in reverse order. For instance, "xyzabczyx"

## Exercise 4

```
1.  Start and end with the same character: \^(.)\w\*\\1\$
2.  Contain a repeated pair of letters: (\w\w).\*\\1
3.  Contain one letter repeated in at least three places: (.)\w\*\\1\w\*\\1
```


