---
title: "Project 2: Data Analysis on Storm and Employee Datasets"
author: "Amish Rasheed"
date: "2024-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(readr)
library(dplyr)
```

# Dataset 1: Historical Storms 2024 Data

In this section, I analyzed the `Historical Storms 2024` data to explore the relationship between storm surge and impact variables, such as economic damage. Additionally, I investigated which states are more frequently affected by named storms.

## Loading the Storm Dataset

```{r}
url <- "https://raw.githubusercontent.com/Amish22/DS607/refs/heads/main/Historical-Storm-Data.csv"
storm_data <- read.csv(url, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(storm_data)
```

## Separating the States into Separate Columns

To analyze the number of named storms by state, we need to split the `Affected Areas` column into separate columns for each state. To achieve this using the `separate()` function from `tidyr`.

```{r}
storm_data <- storm_data %>%
  separate(`Affected.Areas`, into = paste0("Area", 1:7), sep = ", ", fill = "right")
```

## Handling Missing Values

Missing values can impact analysis and visualizations. To address this, replace all `NA` values in the dataset with a placeholder text `"No Data"`.

```{r}
storm_data <- storm_data %>%
  mutate(across(everything(), ~ replace(., is.na(.), "No Data")))
```

## Analysis: Relationship Between Surge Data and Impact Variables

We are particularly interested in understanding whether higher storm surges are associated with greater economic damage. For this analysis, convert the `Damage (bn USD)` and `Observed Storm-Tide` columns to numeric values and compute their correlation.

```{r}
storm_data$Damage..bn.USD. <- as.numeric(gsub("[^0-9.]", "", storm_data$Damage..bn.USD.))
storm_data$Surge <- as.numeric(gsub("[^0-9.]", "", storm_data$Observed.Storm.Tide))

# Checking correlation between surge and damage
correlation <- cor(storm_data$Surge, storm_data$Damage..bn.USD., use = "complete.obs")
correlation
```

**Summary**: A correlation value close to 1 or -1 would indicate a strong linear relationship. In our case, a positive correlation suggests that higher storm surges may be associated with increased economic damage.

## Analyzing the Number of Named Storms by State

Next, to identify which states have experienced the highest number of named storms, reshape the data and count the number of named storms for each state.

```{r}
state_counts <- storm_data %>%
  pivot_longer(cols = starts_with("Area"), values_drop_na = TRUE) %>%
  group_by(value) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

print(state_counts)
```

**Summary**: The table above shows the frequency of named storms for each state, helping us identify the regions that are most frequently impacted.

## Visualization: Relationship Between Surge and Damage

Create a scatter plot to visualize the relationship between observed storm-tide (surge) and economic damage.

```{r}
ggplot(storm_data, aes(x = Surge, y = Damage..bn.USD.)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship between Surge and Damage (in billion USD)",
       x = "Observed Storm-Tide (Surge)",
       y = "Damage (bn USD)")
```

# Dataset 2: Employee Salary Data Analysis

The second dataset focuses on employee salaries across different departments and divisions. Below I tried to understand the average compensation by gender, compensation distribution across departments, and how compensation varies with grade and division.

## Loading the Employee Salary Dataset

```{r}
url <- "https://raw.githubusercontent.com/Amish22/DS607/refs/heads/main/Employee_Salaries_-_2023.csv"
employee_data <- read.csv(url)
head(employee_data)
```

## Data Transformation: Converting Salary Columns to Numeric

To perform meaningful analysis, first convert the `Base_Salary`, `Overtime_Pay`, and `Longevity_Pay` columns to numeric values.

```{r}
employee_data <- employee_data %>%
  mutate(Base_Salary = as.numeric(Base_Salary),
         Overtime_Pay = as.numeric(Overtime_Pay),
         Longevity_Pay = as.numeric(Longevity_Pay))
head(employee_data)
```

## Calculating Total Compensation

Total compensation is calculated by summing up base salary, overtime pay, and longevity pay.

```{r}
employee_data <- employee_data %>%
  mutate(Total_Compensation = Base_Salary + Overtime_Pay + Longevity_Pay)
```

## Analysis: Average Compensation by Gender

Explore whether there are differences in average compensation between genders.

```{r}
gender_summary <- employee_data %>%
  group_by(Gender) %>%
  summarise(Average_Compensation = mean(Total_Compensation, na.rm = TRUE))

ggplot(gender_summary, aes(x = Gender, y = Average_Compensation, fill = Gender)) +
  geom_bar(stat = "identity", width = 0.6) +
  theme_minimal() +
  labs(title = "Average Compensation by Gender",
       x = "Gender",
       y = "Average Total Compensation") +
  scale_fill_manual(values = c("skyblue", "lightcoral"))
```

## Analyzing Compensation by Department

Calculate the average compensation by department and division to identify where salaries are concentrated.

```{r}
department_summary <- employee_data %>%
  group_by(Department, Division) %>%
  summarise(Average_Department_Compensation = mean(Total_Compensation, na.rm = TRUE))
```

## Compensation by Grade and Division

Finally, create a table to view the average compensation by grade across different divisions.

```{r}
compensation_by_grade <- employee_data %>%
  group_by(Division, Grade) %>%
  summarise(Avg_Compensation = mean(Total_Compensation, na.rm = TRUE)) %>%
  pivot_wider(names_from = Grade, values_from = Avg_Compensation)
```

**Summary**: The table above shows the average compensation for each grade across divisions, providing insights into pay distribution across different levels.

```{r}
View(gender_summary)
View(department_summary)
View(compensation_by_grade)
```
