---
title: "Week 5 Assignment"
author: "Amish Rasheed"
date: "2024-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
# Loading required libraries
library(tidyr)
library(dplyr)
library(ggplot2)

# Load the dataset
url <- "https://raw.githubusercontent.com/Amish22/DS607/refs/heads/main/airline-data.csv"
airline_data <- read.csv(url)

head(airline_data)
```

## Data Transformation: Converting to Long Format

The dataset is initially in a wide format, where each row corresponds to a combination of airline and status (on time or delayed), and columns represent counts of flights for five destinations. To facilitate analysis, I reshaped this data into a long format.

```{r}
# Converting data to long format 
tidy_airline_data <- airline_data %>%
  pivot_longer(cols = c("Los.Angeles", "Phoenix", "San.Diego", "San.Francisco", "Seattle"),
               names_to = "Destination",
               values_to = "Count")

# View the tidied data
print(tidy_airline_data)
```

The tidied data now has three primary columns: `Airline`, `Status`, and `Destination`. The `Count` column represents the number of flights for each combination of airline, status, and destination.

## Handling Missing Data

Next, I addressed missing data by replacing empty cells with zeros, ensuring that all combinations have values for accurate analysis.

```{r}
# Replacing missing data with zeros
tidy_airline_data <- tidy_airline_data %>%
  mutate(Count = ifelse(is.na(Count), 0, Count))
```

## Analysis: Total Flights by Status and Destination

I grouped the data by `Airline`, `Status`, and `Destination` to calculate the total number of flights in each category.

```{r}
summary_data <- tidy_airline_data %>%
  group_by(Airline, Status, Destination) %>%
  summarize(Total_Flights = sum(Count), .groups = "drop")

print(summary_data)

# Visualizing Total Flights by Airline, Status, and Destination
ggplot(summary_data, aes(x = Destination, y = Total_Flights, fill = Status)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ Airline) +
  theme_minimal() +
  labs(title = "Total Flights by Status for Each Airline Across Destinations",
       x = "Destination",
       y = "Total Flights",
       fill = "Flight Status") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Calculating Delay Percentages

The delay percentage provides an easy way to compare on-time performance between the two airlines. A lower delay percentage indicates better on-time performance. This metric will be used in the visualization below to highlight differences.

```{r}
# Calculating delay percentages
delay_percentage <- summary_data %>%
  pivot_wider(names_from = Status, values_from = Total_Flights) %>%
  mutate(Delay_Percentage = (delayed / (`on time` + delayed)) * 100) %>%
  select(Airline, Destination, Delay_Percentage)

print(delay_percentage)
```

## Visualizing Delay Percentages Across Destinations

The bar plot provides a side-by-side comparison of delay percentages for ALASKA and AM WEST across the five destinations.

```{r}
ggplot(delay_percentage, aes(x = Destination, y = Delay_Percentage, fill = Airline)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Comparison of Delay Percentages for ALASKA and AM WEST",
       x = "Destination", y = "Delay Percentage (%)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Findings

From the plot, we can see that AM WEST generally has higher delay percentages compared to ALASKA, indicating that it struggles more with on-time performance.

## Identifying Best and Worst Performance Destinations

Next, I identified the destinations with the highest and lowest delay percentages for each airline.

```{r}
# Destination with highest and lowest delay percentage for each airline
highest_delay <- delay_percentage %>%
  group_by(Airline) %>%
  slice_max(Delay_Percentage)

lowest_delay <- delay_percentage %>%
  group_by(Airline) %>%
  slice_min(Delay_Percentage)

highest_delay
lowest_delay
```

### Summary of Best and Worst Performance

The destinations with the highest and lowest delay percentages highlight which routes perform well and which need improvement.

## Statistical Summary of Delay Percentages

To provide more context, I calculated the average and median delay percentages for each airline.

```{r}
# Calculating average and median delay percentages
stat_summary <- delay_percentage %>%
  group_by(Airline) %>%
  summarize(Average_Delay = mean(Delay_Percentage, na.rm = TRUE),
            Median_Delay = median(Delay_Percentage),
            .groups = "drop")

stat_summary

# Transforming for visualization
stat_summary_long <- stat_summary %>%
  pivot_longer(cols = c("Average_Delay", "Median_Delay"),
               names_to = "Statistic",
               values_to = "Percentage")

# Visualizing average and median delays for each airline
ggplot(stat_summary_long, aes(x = Airline, y = Percentage, fill = Statistic)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Average and Median Delay Percentages for ALASKA and AM WEST",
       x = "Airline", y = "Delay Percentage (%)") +
  scale_fill_brewer(palette = "Set2")
```

### Summary

AM WEST has a higher average delay percentage compared to ALASKA. The high average delay indicates that AM WEST experiences more frequent delays. The median delay percentage for AM WEST is also higher than ALASKA, indicating that even typical delays are more common for AM WEST.

### Overall and City-by-City Comparisons

Comparing the overall percentages with the city-by-city performance reveals that AM WEST struggles more consistently across cities, whereas ALASKA maintains relatively better performance in most locations.
